Variance reduction for Markov chains via martingale representations

### 2020-04-15 aap/Revision/vr-mcmc-revision-new-new.tex

In all the sequel, without loss of generality, we assume that $\nabla U(0)=0$ and we set
\begin{equation}
\label{eq:definition-V}
V(x) = 1 + \|x\|^2 \eqsp.
\end{equation}
\begin{lem}
  \label{lem:quadratic_behaviour}
  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ \| x \| \geq K_2$ $\ps{\nabla U(x)}{x} \geq (m/2) \| x \|^2$.
\end{lem}
\begin{proof}
  Using {\bf (H1)} and {\bf (H2)}, we have for any $x \in \rset^d$, $\| x \|\geq K_1$,
  \begin{align*}
    \ps{\nabla U(x)}{x}
    &= \int_{0}^{K_1/\| x \|} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/\| x \|} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
    & \geq m \|x \|^2 \{1- K_1 (1 +L/m)   / \| x \| \}\eqsp,
  \end{align*}
which proves the first statement. The second statement is obvious.
\end{proof}
\begin{lem}
\label{lem:drift}
 Assume that {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
 Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/4$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
 \begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
\end{equation}
with   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = 25K_2^2/8 + d + m$ with $K_2$ from \eqref{eq:H3}, $m$ from {\bf (H2)}.
\end{lem}

\begin{proof} 
$$
Q_{\gamma}V(x) = \int\limits_{\rset^d}V(y)Q_{\gamma}(x,dy) = \int\limits_{\rset^d}\frac{(1+\|y\|^2)}{(2\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{2\gamma}\right)}\,dy =
$$

$$
=1 + \frac{1}{(2\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{2\gamma}\right)}\,dz
$$
Let us first consider the case $x \notin B(0,K_2)$. Note that
$$
\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \|x - \gamma \nabla U(x)\|^2
$$
and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,\gamma I_d)$, it holds $\mathbb{E}\|Z\|^2 = \gamma d$. It remains to notice that due to \eqref{eq:H3},
\[
\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq \left(1 - \gamma m + 2\gamma^2 L^2\right)\|x\|^2
\]
Since $\gamma < \frac{m}{4L}$, we obtain
$$
Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (d + m)
$$
Now let $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
$$
Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + d + 2(1+L\gamma)^2\|x\|^2\right) \leq
$$

>>>>

In all the sequel, without loss of generality, we assume that $\nabla U(0)=0$ and we set
\begin{equation}
\label{eq:definition-V}
V(x) = 1 + \|x\|^2 \eqsp.
\end{equation}
\begin{lem}
  \label{lem:quadratic_behaviour}
  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ \| x \| \geq K_2$ $\ps{\nabla U(x)}{x} \geq (m/2) \| x \|^2$.
\end{lem}
\begin{proof}
  Using {\bf (H1)} and {\bf (H2)}, we have for any $x \in \rset^d$, $\| x \|\geq K_1$,
  \begin{align*}
    \ps{\nabla U(x)}{x}
    &= \int_{0}^{K_1/\| x \|} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/\| x \|} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
    & \geq m \|x \|^2 \{1- K_1 (1 +L/m)   / \| x \| \}\eqsp,
  \end{align*}
which proves the first statement. The second statement is obvious.
\end{proof}
\begin{lem}
\label{lem:drift}
 Assume that {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
 Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/4$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
 \begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
\end{equation}
with   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = 25K_2^2/8 + d + m$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $m$ from {\bf (H2)}.
\end{lem}

\begin{proof} 
Note first that
\[
Q_{\gamma}V(x) = 
1+ \| x - \gamma \nabla U(x) \|^2 + \gamma d
\]
Let us first consider the case $x \notin B(0,K_2)$.  
It remains to notice that, by Lemma~\ref{lem:quadratic_behaviour}, we get
\[
\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq \left(1 - \gamma m + 2\gamma^2 L^2\right)\|x\|^2
\]
Since $\gamma < \frac{m}{4L}$, we obtain
$$
Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (d + m)
$$
Consider now the case $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
$$
Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + d + 2(1+L\gamma)^2\|x\|^2\right) \leq
$$
